---
title: "Lecture 2: Introduction to R programming"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    fig_caption: true
    number_sections: true
---

# FIN 510 setup for this lecture

This lecture assumes that you are running RStudio on an Amazon EC2 instance created from AMI `FIN510-RStudio` with IAM role `FIN510EC2Role`.

On your EC2 instance, create a folder for today's lecture. In the RStudio File window, create a New Folder in your Home directory, and name the folder `fin510`. Inside this folder, create a new subdirectory called `lecture-02`. The path to this directory is `~/fin510/lecture-02`, where the tilde (`~`) character refers to your Home directory.

For today's class, copy the file `lecture-02-tryit.Rmd` from the course S3 bucket, `s3://bigdata-fin510`, to `~/fin510/lecture-02` using the following [AWS Command Line Interface (CLI)](https://aws.amazon.com/cli/#file_commands_anchor) command. This is a Unix command, not an R command, and thus must be passed directly to the operating system. RStudio allows you to execute operating system commands using the built-in shell, found at `Tools` \> `Shell`.

```{bash, eval=FALSE}
# This is a Unix command. Enter via RStudio Shell
aws s3 cp --recursive s3://bigdata-fin510/lecture-02 ~/fin510/lecture-02
```

# Introduction to R and RStudio

## What is R?

R is an open source (free) programming language that is popular among statisticians, quantitative analysts, and data scientists for data analytics. Many powerful data manipulation and analysis tools are available in R through add-on packages that extend base R functionality. As a result, the R programming environment continues to evolve and expand.

## What is RStudio?

RStudio is an integrated development environment (IDE), that facilitates R programming. RStudio includes a code editor, console, file explorer, visualization tools, and many other features that facilitate R programming. In this class, we will use the RStudio Server edition which runs on a remote server with access via a web browser. You can also [install](https://www.rstudio.com/products/rstudio/download/) and run the RStudio Desktop edition on your desktop (Windows, Mac, and Linux). However, many exercises we do in class assume you are using RStudio on a server running within the AWS cloud computing ecosystem.

## Getting help with R and RStudio

Many helpful resources for R programming and RStudio functionality can be found within RStudio itself.

-   You can learn about R functions, data sets, and packages by using the `?` help operator. To learn about function `foo`, type `?foo` in the Console and press Ctrl+Enter. **Try it: what does the `lm` function do?**
-   If you do not know the exact name of the function you are looking for, you can search the R help documentation using the `??` help search operator. To search for the term `foo`, type `??foo` in the Console and press Ctrl+Enter. If your search term contains spaces, contain the phrase in quotes, like `??"foo fighters"`. **Try it: what do you get for searching for the term `model`? How about the phrase `linear model`?**
-   Another excellent way to find help is to use an internet search engine. The ability to find help online is crucial skill for any programmer or data scientist.
-   RStudio has a number of "cheat sheets" that make it easy to learn or use common R packages and RStudio functionality. These cheat sheets can be found [online](https://www.rstudio.com/resources/cheatsheets/) or from within RStudio by going to Help \> Cheatsheets.

## Console vs. Script

The [RStudio Console](https://support.rstudio.com/hc/en-us/articles/200404846-Working-in-the-Console) allows you to enter R commands interactively, much as you would with a high-powered calculator. You can retrieve previous commands using the `↑` (up arrow) and `↓` (down arrow) keys.

RStudio has many keyboard shortcuts to facilitate programming. What does the shortcut `Alt+Shift+K` do?

In this class, we will use [R Markdown](http://rmarkdown.rstudio.com) to develop R code, analyze data, and generate high quality reports. Markdown allows text, code, and results to be woven together in the same document. RStudio provides a quick start and helpful reference at `Help` \> `Markdown Quick Reference`.

To learn more about R Markdown, see tutorials from [RStudio](http://rmarkdown.rstudio.com/lesson-1.html) and [CMU](http://www.stat.cmu.edu/~cshalizi/rmarkdown/), an [R Markdown Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf). For a deeper dive into creating documents using R Markdown, see [*R Markdown: The Definitive Guide*](https://bookdown.org/yihui/rmarkdown/).


# R Coding Fundamentals

## Help within R

RStudio has a "Help" tab that displays R help pages. You can search the documentation by entering the name of a function or keyword into the search bar. You can also access the help pages directly using R commands.

```{r, eval=FALSE}
# help for "typeof", using help()
help(typeof)

# help for "typeof", using ?
?typeof
```

## Vectors, Lists, and Data Types

The fundamental building blocks of data structures in R are are called "vectors." Vectors can store one or more data values and come in two flavors:

1.  **Atomic vectors**, whose elements are of the same type. The most common types are **numeric** (integer or decimal), **character** (strings), and **logical** (true and false). Elements of a single type can be combined into a vector using the `c()` function.
2.  **Lists**, whose elements can be of differing type. In particular, you can have lists that contain other lists. Elements can be combined into a list using the `list()` function.

Given a vector, the function `typeof()` returns its type and `length()` returns the number of elements it contains.

### Numeric vectors

```{r, eval=FALSE}
# A numeric vector of length 1
1.5
typeof(1.5)

# A numeric vector of length >1
c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34)
typeof(c(0, 1, 1, 2, 3, 5, 8, 13, 21, 34))

# The colon operator `:` generates sequence of integers in steps of 1 or -1
1:10
10:1

# Arithmetic Operators
#   +   add
#   -   subtract
#   *   multiply
#   /   divide
#   ^   raise to the power
2^2
```

### Character vectors

```{r, eval=FALSE}
# Character values are delimited by quotation marks

# A character vector of length 1
"Hello World!"
typeof("Hello World!")
length("Hello World!")

# A character vector of length > 1
c("Hello", "World", "!")
typeof(c("Hello", "World", "!"))
length(c("Hello", "World", "!"))
```

### Logical vectors

```{r, eval=FALSE}
# The words TRUE and FALSE denote the two logical constants
TRUE
FALSE
typeof(TRUE)
typeof(FALSE)

# A logical vector of length > 1
c(TRUE, TRUE, TRUE, FALSE)
typeof(c(TRUE, TRUE, TRUE, FALSE))

# Logical operators
#   &   for logical AND 
#   |   for logical OR
#   !   for logical negation (NOT)
TRUE | FALSE
TRUE & FALSE
!TRUE

# Relational operators: 
#   <   for Less than
#   >   for Greater than
#   <=  for Less than or equal to
#   >=  for Greater than or equal to
#   ==  for Equal to
#   !=  for Not equal to
5 > 2
5 >= 2
5 < 2
5 <= 2
5 == 2
5 != 2
```

### Lists

```{r, eval=FALSE}
# Lists can contain elements of any type
list("Molitor", "FIN", 510)

# Lists can even contain other lists
list("Molitor", list("FIN", 510))

# The elements of a list can be named using Name = Value pairs for each element
list(Name = "Molitor", Subject = "FIN", Number = 510)

# names() returns the names of list elements
names(list(Name = "Molitor", Subject = "FIN", Number = 510))
```

## Variable assignment

```{r, eval=FALSE}
# Operators `<-` and `=` assign a value to a variable name
h <- c("Hello", "World", "!")
# or
h = c("Hello", "World", "!")

# `print()` prints the contents of variable
print(h)

# Implicit printing
h

# `rm()` deletes a variable
rm(h)
```

## Subsetting vectors and lists

Vectors and lists can be subset using single brackets `[ ]`, double brackets `[[ ]]`, and dollar sign `$` operators. We will illustrate how these operators work on the following vector and list.

```{r}
# Define an atomic vector
x_vector = c(2, 4, 6, 8)

# Define a named list
x_list = list(Name = c("David", "Molitor"), Subject = "FIN", Number = 510)
```

### List/vector subsetting (multiple elements)

Use single brackets `[` to select one or more elements of a vector or list.

```{r, eval=FALSE}
# Positive integers select elements in given positions
x_vector[c(2, 3)]

# Negative integers drop elements in given positions
x_vector[c(-1, -4)]

# Logical vectors select elements with a corresponding value of TRUE
x_vector[c(FALSE, TRUE, FALSE, TRUE)]

# For named lists, you can select elements by name
x_list[c("Subject", "Number")]

# When subsetting a list, `[` always returns a list containing the selected elements
x_list[2]
typeof(x_list[2])
```

### Element extraction (single element)

Use double brackets `[[` or the dollar sign `$` to extract a single element from a vector or list.

```{r, eval=FALSE}
# Positive integer extracts element in given position
x_vector[[2]]

# For named lists, can extract an element by name
x_list[["Subject"]]

# Extraction from a list returns the selected element itself, not a list containing the element
x_list[[2]]
typeof(x_list[[2]])
```

## Conditional statements (if, if/else)

```{r, eval=FALSE}
# If
if (TRUE) {
  print("Condition is TRUE!")
}

# If/Else
if (TRUE) {
  print("Condition is TRUE!")
} else {
  print("Condition is FALSE!")
}
```

## Loops (for)

```{r, eval=FALSE}
for (item in vector) {
  print(item)
}
```

## Functions

```{r, eval=FALSE}
# Create a function that adds two numbers
f <- function(a, b) { 
  a + b
}

# Invoke the function
f(a = 2, b = 1)
f(2, 1)
f()

# Redefine function with default values for arguments
f <- function(a = 0, b = 0) { 
  a + b
}
f(a = 2, b = 1)
f(2, 1)
f()
```

## Data frames

Data frames are the most important object in R for data analytics. A data frame is a rectangular data structure, where each row is an observation and each column is a variable that describes some characteristic of the observation.

Technically, data frames are just named lists with a special structure:

-   Each item in the list is a vector
    -   These vectors are the columns, also called "variables," in the data frame

-   All vectors in the list must have equal length
    -   This gives data frames a rectangular structure

-   Columns have names (variable names)


### Create a data frame from scratch

You can create a data frame from scratch, using the `data.frame()` function.

```{r, results = "hide"}
# People and birth country vectors, each of length 2
person <- c("Xiao Wang", "Sri Jena")
country <- c("China", "India")

# Combine into a data frame
X <- data.frame(Name = person, Birth_Country = country)

# X is an object with class “data.frame”
class(X)
```

Because data frames are lists, you can subset them using the techniques discussed above.

```{r, eval=FALSE}
# Extract the first row of the data frame
X[1,]

# Extract the first column of the data frame
X[,1]
X[[1]]

# Data on people named "Sri Jena"
X[X$Name == "Sri Jena"]

# Extract the vector of birth countries
X$Birth_Country
```

## Read external data into a data frame

Usually, large data sets come stored in external databases or text data like comma separated values (csv). The `tidyverse` package contains R functions that can read most common types of flat file data directly into a data frame.

```{r, eval=FALSE}
# Load package
library(tidyverse)

# CSV "Comma Separated Values" Format
path <- "https://raw.githubusercontent.com/googletrends/data/master/20160819_OlympicSportsByCountries.csv"
trends.csv <- read_csv(path, skip=2, n_max=100)

# Display the first few rows
head(trends.csv)
trends.csv[1:3,]

# Display countries interested in Archery
trends.csv[trends.csv$Sport == "Archery", ]
```
