---
title: "Lecture 16: Neural Nets"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

<font size="5" style="color:blue"> first example: predicting customers’ preference (binary outcome)  </font> 

## 1 create a data frame 
```{r}
# load the data
df <- read.csv("TinyData.csv")

# create two outcome dummies  
df$Like <- df$Preference=="like"
df$Dislike <- df$Preference=="dislike"

df
```

## 2 neural network with a single hidden layer 
```{r}
library(neuralnet)

set.seed(1)
nn1<- neuralnet(Like + Dislike ~ x2 + x1, data = df, linear.output = FALSE, hidden = 3)
# hidden: a vector of integers specifying the number of hidden neurons in each layer
# linear.output=TRUE for regression and linear.output=FALSE for classification

# display weights
nn1$weights

# plot the network 
# rep="best": the repetition of the neural network with the smallest error will be plotted
plot(nn1, rep = "best")

# display the activation function
nn1$act.fct
```

## 3 making predictions

### 3.1 predictions for a data set  
```{r}
# create a data frame 
data.frame(df$x2, df$x1)

predict <- compute(nn1, data.frame(df$x2, df$x1))
```

### 3.2 predicted probabilities (almost)
```{r}
predict$net.result

# predicted probabilities of like 
predict$net.result[,1]

# predicted probabilities of dislike 
predict$net.result[,2]
```

### 3.3 predicted classes
```{r}
predicted.class <- ifelse(predict$net.result[,2]>predict$net.result[,1], "dislike", "like")
predicted.class
```

### 3.4 confusion matrix 
```{r}
# predicted classes are character values  
predicted.class

# actual classes are character values 
df$Preference

library(caret)
confusionMatrix(as.factor(predicted.class), as.factor(df$Preference),positive="like")
```
<font size="5" style="color:blue"> second example: predicting prices of used Toyota Corollas </font> 

## 4 create a data frame 

### 4.1 load the data 
```{r}
toyota.df <- read.csv("ToyotaCorolla.csv")
head(toyota.df)
```

### 4.2 transforming Fuel_Type into 2 dummies 
```{r}
# Fuel_Type has 3 levels   
toyota.df$Fuel_Type_CNG <- ifelse(toyota.df$Fuel_Type == "CNG",1,0)
toyota.df$Fuel_Type_Diesel <- ifelse(toyota.df$Fuel_Type == "Diesel",1,0)
```

### 4.3 data partition
```{r}
set.seed(1)

# total number of rows 
dim(toyota.df)[1]

# number of rows to select for the training set 
0.6*dim(toyota.df)[1]

# row indexes of the training set
train.index <- sample(c(1:dim(toyota.df)[1]), 0.6*dim(toyota.df)[1])  
head(train.index)

# training set
train.df <- toyota.df[train.index, ]
head(train.df)

# test set 
test.df <- toyota.df[-train.index, ]
head(test.df)
```

### 4.4 scale the numerical predictors and the outcome variable 
```{r}
# maximum price in the training set 
max(train.df$Price)

# minimum price in the training set 
min(train.df$Price)

# training set 
head(train.df$Price)

# normalized price for the first record 
(9250-min(train.df$Price))/(max(train.df$Price)-min(train.df$Price))

# estimate the transformation 
library(caret)
norm.values <- preProcess(train.df, method="range")
# method="range": scale the data to the interval between zero and one

# normalize the numerical predictors and the outcome variable in the training set 
train.norm.df <- predict(norm.values, train.df)
# Price, Age_08_04, KM, HP, Doors, Quarterly_Tax, Guarantee_Period are normalized 
head(train.norm.df)

# normalize the numerical predictors and the outcome variable in the test set 
test.norm.df <- predict(norm.values, test.df)
head(test.norm.df)
```

## 5 single hidden layer with 2 nodes 

### 5.1 fit a neural network on the training set 
```{r}
# set the seed for reproducing the result 
set.seed(1)

nn2 <- neuralnet(Price ~ Age_08_04+KM+Fuel_Type_CNG+Fuel_Type_Diesel+HP+Automatic+Doors+Quarterly_Tax+Mfr_Guarantee
                +Guarantee_Period+Airco+Automatic_airco+CD_Player+Powered_Windows+Sport_Model+Tow_Bar, 
                data = train.norm.df, linear.output = TRUE, hidden = 2)
# hidden: a vector of integers specifying the number of hidden neurons in each layer
# linear.output=TRUE for regression and linear.output=FALSE for classification
```

### 5.2 plot the network
```{r}
# rep="best": the repetition of the neural network with the smallest error will be plotted
plot(nn2, rep = "best")
```

### 5.3 making predictions for records in the test set 
```{r}
predict.nn2 <- compute(nn2, test.norm.df)

# predicted prices (normalized)
head(predict.nn2$net.result)
```

### 5.4 prediction error on the test set
```{r}
mean((test.norm.df$Price-predict.nn2$net.result)^2)
```
