---
title: "Lecture 4: Data Exploration"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 1 create a tibble 
```{r}
# load the data 
library(tidyverse)
df <- read_csv("exe.csv")

# first six rows 
head(df)

# number of rows and columns 
dim(df)
nrow(df)
ncol(df)

# column names 
names(df)

# variable type 
str(df)
```

## 2 summary statistics 
```{r}
# minimum 
min(df$SALARY)

# maximum
max(df$SALARY)

# mean
mean(df$SALARY)

# median
median(df$SALARY)

# first and third quantile 
quantile(df$SALARY, 0.25)
quantile(df$SALARY, 0.75)
quantile(df$SALARY, c(0.25, 0.75))

# summary 
summary(df$SALARY)

# distinct 
n_distinct(df$TICKER)

# frequency count 
table(df$TICKER)
count(df,TICKER)
```

## 3 counts and proportion of logical values  

### 3.1 counts of logical values 
```{r}
# SALARY values 
head(df$SALARY)
# whether SALARY>1000
head(df$SALARY>1000)
# number of rows with SALARY >1000
sum(df$SALARY>1000)

# YEAR values
head(df$YEAR)
# whether YEAR=2015 
head(df$YEAR==2015)
# number of rows with YEAR = 2015 
sum(df$YEAR==2015)

# CFOANN values
head(df$CFOANN)
# whether CFOANN is missing 
head(is.na(df$CFOANN))
# number of rows with CFOANN=CFO (not missing)
sum(!is.na(df$CFOANN))

# dealing with missing values 
sum(df$CFOANN=="CFO",na.rm=TRUE)
# wrong
sum(df$CFOANN=="CFO")
```

### 3.2 proportions of logical values 
```{r}
# proportion of rows with SALARY >1000
mean(df$SALARY>1000)

# proportion of rows with YEAR = 2015 
mean(df$YEAR==2015)

# proportion of rows with CFOANN=CFO
mean(!is.na(df$CFOANN))
# wrong 
mean(df$CFOANN=="CFO",na.rm=T)
```

## 4 summarize()

```{r}
# average and maximum SALARY, number of observations, and number of unique TICKER 
summarize(df, 
          SALARY_mean=mean(SALARY),
          SALARY_max=max(SALARY),
          n_rows=n(),
          n_firms=n_distinct(TICKER))

# counts of CEOs and CFOs
summarize(df,
          CEOANN_count=sum(CEOANN=="CEO",na.rm=T),
          CFOANN_count=sum(CFOANN=="CFO",na.rm=T))

# average and maximum SALARY, and the number of observations by TICKER and YEAR 
df_by_TICKER_YEAR <- group_by(df,TICKER, YEAR)
summarize(df_by_TICKER_YEAR, 
          SALARY_mean=mean(SALARY),
          SALARY_max=max(SALARY),
          n_rows=n(),.groups="drop")
```

## 5 combine multiple operations with the pipe

### 5.1 salary plus bonus for CEOs in 2019
```{r}
df %>% 
  filter(CEOANN=="CEO" & YEAR==2019) %>% 
  mutate(SALARY_BONUS=SALARY+BONUS) %>% 
  select(TICKER,EXEC_FULLNAME,SALARY,BONUS,SALARY_BONUS)
```

### 5.2 average salary by firm
```{r}
df %>% 
  group_by(TICKER) %>% 
  summarize(SALARY_mean=mean(SALARY)) %>% 
  arrange(desc(SALARY_mean))
```

### 5.3 firms with average salary more than 900k
```{r}
df %>% 
  group_by(TICKER) %>% 
  summarize(SALARY_mean=mean(SALARY)) %>% 
  filter(SALARY_mean>900) %>%
  arrange(desc(SALARY_mean)) 
```

### 5.4 the highest paid executive's salary by firm 
```{r}
df %>% 
  group_by(TICKER) %>% 
  arrange(desc(TDC1)) %>% 
  summarize(SALARY_highest_TDC1=first(SALARY))

df %>% 
  group_by(TICKER) %>% 
  arrange(TDC1) %>% 
  summarize(SALARY_highest_TDC1=last(SALARY))
```

### 5.5 highest salary plus bonus by firm 
```{r}
df %>% 
  mutate(SALARY_BONUS=SALARY+BONUS) %>% 
  group_by(TICKER) %>% 
  summarize(SALARY_BONUS_max=max(SALARY_BONUS))
```

### 5.6 average salary by gender 
```{r}
df %>% 
  group_by(GENDER) %>% 
  summarize(SALARY_mean=mean(SALARY),
            n_rows=n()) 
```

### 5.7 CEOs or CFOs average salary by gender 
```{r}
df %>% 
  filter(CEOANN=="CEO" | CFOANN=="CFO") %>% 
  group_by(GENDER) %>% 
  summarize(SALARY_mean=mean(SALARY))
```

### 5.8 cumulative salaries by firm in 2015
```{r}
df %>% 
  filter(YEAR==2015) %>% 
  group_by(TICKER) %>% 
  summarize(SALARY, 
            SALARY_cumsum=cumsum(SALARY),
            SALARY_sum=sum(SALARY), .groups="drop")
```

