---
title: "Lecture 7: Linear Regression"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

### not display scientific notation
```{r}
# scipen: a penalty to be applied when deciding to print numeric values in fixed or exponential notation. 
# positive values bias towards fixed and negative towards scientific notation
options(scipen=999)
```

## create a data frame 
```{r}
# read data from a csv file 
housing.df <- read.csv("ameshousing.csv")

# first six rows 
head(housing.df)

# column names 
names(housing.df)

# summary statistics 
summary(housing.df)
```

## 1 intercept only linear model  
```{r}
# intercept only model where the outcome is Sale Price
lm1 <- lm(SalePrice ~ 1, data = housing.df)
summary(lm1)

# coefficient is the mean 
mean(housing.df$SalePrice)
```

## 2 model with a numeric (continuous) predictor 
```{r}
# linear regression of Sale Price on above ground living area 
lm2 <- lm(SalePrice ~ Gr_Liv_Area, data = housing.df)
summary(lm2)

# predicted values
head(lm2$fitted.values)

# plot the actual and predicted values 
library(ggplot2)
ggplot(housing.df, aes(x = Gr_Liv_Area, y = SalePrice)) + geom_point(color="grey",size=2) + 
  geom_line(aes(x = Gr_Liv_Area, y=lm2$fitted.values), color="blue")
```

the parameter estimate for the intercept is 18583.434;
the parameter estimate for the slope of Gr_Liv_Area is 105.189;
the regression equation is Sale Price=18583.434+105.189*Gr_Liv_Area;
the model indicates that each additional square foot of Gr_Liv_Area is associated with an approximately $105.189  higher sale price.

## 3 model with a numeric (discrete) predictor  
```{r}
# linear regression of Sale Price on the number of Bedroom above the ground 
lm3 <- lm(SalePrice ~ Bedroom_AbvGr, data = housing.df)
summary(lm3)
```

## 4 model with an ordinal categorical predictor with character values   

### 4.1 character values are automatically converted to dummies in regressions 
```{r}
# four levels (character values)
table(as.factor(housing.df$Heating_QC))

# character values are automatically converted to dummies (Heating_QCFa, Heating_QCGd, Heating_QCTA) in regressions 
lm4a <- lm(SalePrice ~ Heating_QC, data = housing.df)
# compared to Ex (the first level)
summary(lm4a)

# mean of Sale Price for each value in Heating_QC
tapply(housing.df$SalePrice, housing.df$Heating_QC, mean)
# Fa - Ex
97119-154919
# Gd - Ex
130844-154919
# TA - Ex
130574-154919

# plot the mean of Sale Price for each value in Heating_QC
ggplot(housing.df, aes(x=Heating_QC, y=SalePrice)) + 
  geom_bar(stat = "summary", fun = "mean")
```

### 4.2 convert Heating_QC to a categorical variable first 
```{r}
lm4b <- lm(SalePrice ~ as.factor(Heating_QC), data = housing.df)
summary(lm4b)
```

### 4.3 create a design matrix by expanding factors to a set of dummies variables  
```{r}
# 0: without intercept 
head(model.matrix(~ 0+Heating_QC, data = housing.df))

# turn the matrix into a data frame 
Heating_QC_dummy <- as.data.frame(model.matrix(~ 0+Heating_QC, data = housing.df))
head(Heating_QC_dummy)

# four levels need 3 dummies 
head(Heating_QC_dummy[,-1])

# combine the data frames 
housing.df <- cbind(housing.df, Heating_QC_dummy[,-1])
head(housing.df)

# fit a linear regression of Sale Price on three dummies 
lm4c <- lm(SalePrice ~ Heating_QCFa+Heating_QCGd+Heating_QCTA, data = housing.df)
summary(lm4c)
```
## 5 model with a nominal categorical predictor with numeric values 

### 5.1 numeric values will not be automatically converted to dummies
```{r}
# four levels (numeric values)
table(as.factor(housing.df$Season_Sold))

# numeric values will not be automatically converted to dummies
lm5a <- lm(SalePrice ~ Season_Sold, data = housing.df)
# wrong result 
summary(lm5a)
```

### 5.2 convert Season_Sold to a categorical variable first
```{r}
lm5b <- lm(SalePrice ~ as.factor(Season_Sold), data = housing.df)
summary(lm5b)

# mean of Sale Price for each value in Season_Sold
tapply(housing.df$SalePrice, housing.df$Season_Sold, mean)
# 2 - 1 (spring - winter)
140879-120917
# 3-1 (summer - winter)
137410-120917
# 4-1 (fall - winter)
144977-120917

# plot the mean of Sale Price for each value in Season_Sold_cat
ggplot(housing.df,aes(x=Season_Sold, y=SalePrice)) + 
  geom_bar(stat = "summary", fun = "mean")
```

## 6 model with numeric and categorical predictors 
```{r}
# linear regression of Sale Price on Gr_Liv_Area and Heating_QC
lm6 <- lm(SalePrice ~ Gr_Liv_Area+as.factor(Heating_QC), data = housing.df)
summary(lm6)
```

## 7 non-linear model with polynomial predictors  

### 7.1 use poly()
```{r}
lm7a <- lm(SalePrice ~ poly(Gr_Liv_Area, 2, raw = TRUE), data = housing.df)
# raw polynomial terms
summary(lm7a)

# predicted values 
head(lm7a$fitted.values)

# plot the actual and predicted values 
ggplot(housing.df, aes(x = Gr_Liv_Area, y = SalePrice)) + geom_point(color="grey",size=2) + 
  geom_line(aes(x = Gr_Liv_Area, y=lm7a$fitted.values), color="blue")
```

### 7.2 create a new variable 
```{r}
# square Gr_Liv_Area
housing.df$Gr_Liv_Area_sqr <- housing.df$Gr_Liv_Area^2
head(housing.df$Gr_Liv_Area)
head(housing.df$Gr_Liv_Area_sqr)

# regression of Sale Price on Gr_Liv_Area and Gr_Liv_Area_sqr
lm7b <- lm(SalePrice ~ Gr_Liv_Area+Gr_Liv_Area_sqr, data = housing.df)
summary(lm7b)
```
