---
title: "Lecture 8: Logistic Regression"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

### not display scientific notation
```{r}
# scipen: a penalty to be applied when deciding to print numeric values in fixed or exponential notation. 
# positive values bias towards fixed and negative towards scientific notation
options(scipen=999)
```

## 1 creating a data frame 

### 1.1 data pre-processing 
```{r}
bank.df <- read.csv("UniversalBank.csv")
head(bank.df)

# drop ID and zip code columns
bank.df <- bank.df[ , -c(1, 5)]  
head(bank.df)

# numeric values
head(bank.df$Education)

# as.factor() coerces its argument to a factor 
head(as.factor(bank.df$Education))

# convert Education from a numeric variable to a categorical or factor variable 
bank.df$Education <- factor(bank.df$Education, levels = c(1, 2, 3), 
                            labels = c("Undergrad", "Graduate", "Advanced/Professional"))
# levels: unique values in the vector of data 
# labels: character vector of labels for the levels 
# logistic regression will automatically create two dummy variables from the factor's three levels 

head(bank.df$Education)

# Education is a categorical or factor variable
str(bank.df)
```

### 1.2 partition data  
```{r}
# the total number of rows and columns 
dim(bank.df)

# the total number of rows  
dim(bank.df)[1]

# sample size of the training set 
dim(bank.df)[1]*0.6

# set seed for reproducing the partition 
set.seed(2)

# take a sample of row numbers 
train.index <- sample(c(1:dim(bank.df)[1]), dim(bank.df)[1]*0.6)  
head(train.index)

# subset the data frame by extracting rows 
train.df <- bank.df[train.index, ]
head(train.df)

test.df <- bank.df[-train.index, ]
head(test.df)
```

## 2 fitting a logistic regression model with a single predictor 

### 2.1 use glm() with family = "binomial" to fit a logistic regression
```{r}
# glm: general linear model
logit.reg.income <- glm(Personal.Loan ~ Income, data = train.df, family = "binomial") 
summary(logit.reg.income)
```

### 2.2 predict for a new customer with $100,000K income 

#### calculate the probability manually 
```{r}
# estimated coefficients
logit.reg.income$coefficients

# logit 
1*(-6.27275)+100*0.03840

# or matrix multiplication 
logit <- c(1, 100) %*% logit.reg.income$coefficients
logit

# probability 
prob <- 1/(1+exp(-logit))
prob
```

#### use the predict function (input should be a data frame)
```{r}
# create a data frame 
new <- data.frame(Income=100)
new

# type = "response" computes the predicted probability 
predict(logit.reg.income, new, type = "response")

# type = "link" computes logit(p)
predict(logit.reg.income, new, type = "link")
```

### 2.3 generate predicted probabilities
```{r}
logit.reg.income.pred <- predict(logit.reg.income, test.df, type = "response")
head(logit.reg.income.pred)
```

### 2.4 create a confusion matrix
```{r}
# predicted classes (numeric)
head(ifelse(logit.reg.income.pred > 0.5, 1, 0))

# actual classes (numeric)
head(test.df$Personal.Loan)

# predicted classes (factor)
head(as.factor(ifelse(logit.reg.income.pred > 0.5, 1, 0)))

# actual classes (factor)
head(as.factor(test.df$Personal.Loan))

# load the package 
library(caret)

# confusion matrix 
confusionMatrix(as.factor(ifelse(logit.reg.income.pred > 0.5, 1, 0)), as.factor(test.df$Personal.Loan),positive = "1")

# accuracy rate
(1746+65) / (1746+123+66+65) 
```

## 3 logistic regression model with 12 predictors 

### 3.1 use glm() with family = "binomial" to fit a logistic regression
```{r}
logit.reg <- glm(Personal.Loan ~ ., data = train.df, family = "binomial") 
summary(logit.reg)
```
customers who are high earners, have a large family , and spend a lot of money on credit card tend more to accept the offer; 
having graduate or professional education and holding a CD account are associated with higher probabilities of accepting the loan offer;
using online banking and owning a Universal Bank credit card are associated with lower acceptance rates.


### 3.2 compute predicted probabilities 
```{r}
logit.reg.pred <- predict(logit.reg, test.df, type = "response")
head(logit.reg.pred)
```
the first six customers are classified as nonacceptors (0) becuase the probability of accepting the offer is lower than the cutoff of 0.5

### 3.3 create confusion matrix 
```{r}
# predicted classes (numeric)
head(ifelse(logit.reg.pred > 0.5, 1, 0))

# actual classes (numeric)
head(test.df$Personal.Loan)

# predicted classes (factor)
head(as.factor(ifelse(logit.reg.pred > 0.5, 1, 0)))

# actual classes (factor)
head(as.factor(test.df$Personal.Loan))

# confusion matrix  
confusionMatrix(as.factor(ifelse(logit.reg.pred > 0.5, 1, 0)), as.factor(test.df$Personal.Loan),positive = "1")

# accuracy rate 
(1794+123) / (1794+65+18+123) 
```

### 3.4 create a lift chart  

#### constructs a gains table
```{r}
head(test.df$Personal.Loan)
head(logit.reg.pred)

# constructs a gains table to evaluate the performance 
library(gains)
# inputs are a vector of actual responses and a vector of predicted probabilities 
gain <- gains(test.df$Personal.Loan, logit.reg.pred, groups=10)

# cumulative percentage of acceptors 
gain$cume.pct.of.total

# total number of acceptors 
sum(test.df$Personal.Loan)

# cumulative number of acceptors 
gain$cume.pct.of.total*sum(test.df$Personal.Loan)

# y axis values 
c(0,gain$cume.pct.of.total*sum(test.df$Personal.Loan))

# cumulative number of customers 
gain$cume.obs

# x axis values 
c(0,gain$cume.obs)
```

#### plot a lift chart
```{r}
plot(c(0,gain$cume.pct.of.total*sum(test.df$Personal.Loan))~c(0,gain$cume.obs), 
     xlab="cumulative number of customers", ylab="cumulative number of acceptors", type="l")

# y axis values 
c(0,sum(test.df$Personal.Loan))

dim(test.df)[1]

# x axis values 
c(0, dim(test.df)[1])

# add a baseline curve 
lines(c(0,sum(test.df$Personal.Loan))~c(0, dim(test.df)[1]))
```


