---
title: "Lecture 3: Data Wrangling"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: simplex
    number_sections: false
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## 1 create and subset a data frame

### create a data frame 
```{r}
manager <- c("Timothy D. Cook", "D. Bruce Sewell")
salary_in_thousand <- c(2000, 1000)
X <- data.frame(EXEC_FULLNAME = manager, SALARY = salary_in_thousand)
X
```

### select a column as a vector 
```{r}
X$EXEC_FULLNAME
X[['EXEC_FULLNAME']]
X[[1]]
X[,'EXEC_FULLNAME']
X[,1]
```

### select a column as a data frame 
```{r}
X['EXEC_FULLNAME']
X[1]
```

## 2 create and subset a tibble

### create a tibble 
```{r}
library(tidyverse)
Y <- tibble(EXEC_FULLNAME = manager, SALARY = salary_in_thousand)
Y
```

### select a column as a vector 
```{r}
Y$EXEC_FULLNAME
Y[['EXEC_FULLNAME']]
Y[[1]]
```

### select a column as a tibble 
```{r}
Y['EXEC_FULLNAME']
Y[1]

# Matrix subsetting does NOT simplify output if only one column is selected
Y[,'EXEC_FULLNAME']
Y[,1]
```

## 3 join data 
```{r}
# create a tibble 
Z <- tibble(EXEC_FULLNAME =c("Timothy D. Cook", "Luca Maestri"), BONUS = c(0,0))
Z

# left join 
left_join(Y, Z, by = "EXEC_FULLNAME")

# right join 
right_join(Y, Z, by = "EXEC_FULLNAME")

# inner join 
inner_join(Y, Z, by = "EXEC_FULLNAME")

# full join 
full_join(Y, Z, by = "EXEC_FULLNAME")
```

## 4 create a tibble 
```{r}
# load the data 
df <- read_csv("manager.csv")

# first six rows 
head(df)

# number of rows and columns 
dim(df)
nrow(df)
ncol(df)

# column names 
names(df)

# variable types 
str(df)
```

## 5 subset data 

### select one column 
```{r, results = "hide"}
# as a vector 
df$EXEC_FULLNAME
df[['EXEC_FULLNAME']]
df[[7]]

# as a tibble 
df['EXEC_FULLNAME']
df[7]
df[,'EXEC_FULLNAME']
df[,7]
select(df,EXEC_FULLNAME)
```

### select two columns 
```{r, results = "hide"}
df[,c("EXEC_FULLNAME","SALARY")]
df[,c(7,9)]
select(df,EXEC_FULLNAME,SALARY)
```

### select five rows and two columns
```{r}
df[c(1:5),c("EXEC_FULLNAME","SALARY")]
df[c(1:5),c(7,9)]
```

## 6 filter rows  

### filter rows 
```{r}
# executives of AAPL in 2015
df$YEAR==2015 & df$TICKER=="AAPL"
df[df$YEAR==2015 & df$TICKER=="AAPL",]
filter(df,YEAR==2015, TICKER=="AAPL")

# executives with the minimum salary
min(df$SALARY)
df$SALARY==min(df$SALARY)
df[df$SALARY==min(df$SALARY),]
filter(df,SALARY==min(SALARY))
```

### filter rows and select columns 
```{r}
# name and salary of executives of AAPL in 2015
df[df$YEAR==2015 & df$TICKER=="AAPL",c("EXEC_FULLNAME","SALARY")]
```

## 7 arrange rows 
```{r}
# ascending order of TDC1 
arrange(df,TDC1)

# descending order of TDC1 
arrange(df,desc(TDC1))

# ascending order of TICKER and YEAR, and descending order of TDC1 
arrange(df,TICKER, YEAR, desc(TDC1))

# does not modify df in place 
df

# modify df in place 
df<- arrange(df,TICKER, YEAR, desc(TDC1))
df
```

## 8 add a column 

### 8.1 salary plus bonus 
```{r}
df$SALRAY_BONUS<-df$SALARY+df$BONUS
head(df)
```

### 8.2 salary bin numbers  
```{r}
# identify the thresholds of 5 equal width bins 
bins <- seq(min(df$SALARY), max(df$SALARY), (max(df$SALARY) - min(df$SALARY))/5)
# 6 thresholds 
bins

# bin SALARY  
df$SALARY_bin <- .bincode(df$SALARY, bins, include.lowest = TRUE)

# frequency counts of unique values
table(df$SALARY_bin)
```

### 8.3 salary intervals 
```{r}
df$SALARY_cut <- cut(df$SALARY, 5)

# frequency counts of unique values
table(df$SALARY_cut)

# not display scientific notation 
format(1.22e+03, scientific = FALSE)
```

### 8.4 salary and bonus in dollars 
```{r}
# does not modify df in place 
mutate(df,SALRAY_1000=SALARY*1000,BONUS_1000=BONUS*1000)
head(df)
```

```{r}
# modify df in place 
df <- mutate(df,SALRAY_1000=SALARY*1000,BONUS_1000=BONUS*1000)
head(df)
```

```{r}
# drop a column
df$SALRAY_BONUS<-NULL
head(df)
```
